Topic modeling bla bla

### 2017-05-31 thesis_report/draft_thesis.tex
Fixes: lines & LDA description

\begin{document}

\maketitle
\tableofcontents

@@

\subsection{Generative process}

\begin{enumerate}
  \item Fix a multinomial distribution with parameter vector $\phi$ of length $V$
  \item for each word in the document:
  \begin{description}
  \item draw a word $w$ according to $\phi$
  \end{description}
\end{enumerate}

Above, step 1 sets up the probability distributions that are then used in step 2 to produce the observed training data. 
A single multinomial distribution can only represent a category of closely related documents. 
For a collection of documents of multiple categories/topics, a simple
generative process is

\begin{enumerate}
\item Fix a multinomial $\alpha$ over categories $1$ to $K$\\
for category number $1$ to category number $K$:
\begin{description}
\item   fix a multinomial with parameter vector $\phi_k$
\end{description}

\item for document number $1$ to document number $M$:
\begin{description}
\item draw a category $z$ according to $\alpha$
\item for each word in the document:
  	\begin{description}
  	\item	draw a word $w$ according to $\phi_z$
  	\end{description}
\end{description}
\end{enumerate}

Note that $z$ is an integer between $1$ and $K$. 
For each document, the value of $z$ is hidden, meaning that it exists conceptually, but it is never known, not even for training data. 
The generative process above gives the following global probability distribution

\begin{equation}
f(x) = \sum\limits_{k=1}^K \alpha_k f(x;\phi_k)
\end{equation}

where $x$ is a document and $\phi_k$ is the parameter vector of the $k$-th multinomial. 
In general, $x$ could be a data point of any type and $\phi_k$ could be the parameters of any appropriate distribution. 
$K$ is called the number of components in the mixture model. 
For each $k$, $f (x; \theta_k)$ is the distribution of component number $k$. 
The scalar $\alpha_k$ is the proportion of component number $k$. 
A distribution like this is called a mixture distribution. 
In general, the components can be probability density functions, for example Gaussians, or probability mass functions, for example multinomials. 
We will see later how to do maximum likelihood estimation for mixture distributions, using a technique called expectation-maximization.

@@

LDA is a topic model that was first presented as a graphical model for topic discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. 
The basic idea is that documents exhibit multiple topics. 
LDA is a generative statistical model that allows sets of observations to be explained by unobserved groups that explain why some parts of the data are similar. 
For example, if observations are words collected into documents, it posits that each document is a mixture of a small number of topics and that each word's occurrence is attributable to one of the document's topics. 
This is similar to PLSI, except that in LDA the topic distribution is assumed to have a Dirichlet prior (see 2.x.x). 
In practice, this results in more reasonable mixtures of topics in a document. 
The reduced document description result in PLSI is a vector where each element describes the mixing proportion for a topic. 
A limitation with PLSI is that there is no generative (see 2.x) model for these proportions, making it difficult to handle unseen documents. 
The Latent Dirichlet Allocation model tries to solve this limitation by setting a Dirichlet prior on the topic distribution.

@@

This statistical model reflects the intuition that documents exhibit multiple topics. 
Each document exhibits the topics with different proportion (step 1); 
each word in each document is drawn from one of the topics (step 2b), where the selected topic is chosen from the per-document distribution over topics (step 2a). 
All the documents in the corpus share the same set of topics, but each document exhibits those topics with different proportion - that’s the distinguishing characteristic of LDA.
A more detailed explanation of the model follows in section 2.x.x.

@@

@@

In LDA, each document is represented by a mixture of the topics, with weight θ2w for topic z in document w. 
These weight have a Dirichlet prior, parameterized by a hyper-parameter α. 
Each topic is a probability distribution over the vocabulary, with probability βwz for word w in topic z.

@@

bla

@@

bla bla

>>>>

\usepackage{amsfonts}
\usepackage{graphicx}

\begin{document}
\maketitle
\tableofcontents

@@

@@

LDA is a topic model that was first presented as a graphical model for topic discovery 
by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic idea is that 
documents exhibit multiple topics. LDA is a generative statistical model that allows 
sets of observations to be explained by unobserved groups that explain why some parts 
of the data are similar. For example, if observations are words collected into 
documents, it posits that each document is a mixture of a small number of topics and 
that each word's occurrence is attributable to one of the document's topics. This is 
similar to PLSI, except that in LDA the topic distribution is assumed to have a 
Dirichlet prior (see 2.3.3.1). In practice, this results in more reasonable mixtures 
of topics in a document. The reduced document description result in PLSI is a vector 
where each element describes the mixing proportion for a topic. A limitation with 
PLSI is that there is no generative model for these proportions, making it difficult 
to handle unseen documents. The Latent Dirichlet Allocation model tries to solve 
this limitation by setting a Dirichlet prior on the topic distribution.

@@

This statistical model reflects the intuition that documents exhibit multiple topics. 
Each document exhibits the topics with different proportion (step 1); each word in 
each document is drawn from one of the topics (step 2b), where the selected topic is 
chosen from the per-document distribution over topics (step 2a). All the documents 
in the corpus share the same set of topics, but each document exhibits those topics 
with different proportion - that’s the distinguishing characteristic of LDA. A more 
detailed explanation of the model follows in section 2.3.3.2.

@@

\begin{itemize}
\item Start out with an empty urn.
\item Randomly pick a colored ball and place it in the urn.
\item Then choose one option:
\begin{enumerate}
\item Randomly pick a colored ball and place it in the urn.
\item Randomly remove a colored ball from the urn, then put it back with another 
ball of the same color.
\end{enumerate}
\end{itemize}

As the number of balls in the urn increase, the probability of picking a new color 
decreases. The proportion of balls in the urn after an infinite amount of draws is 
a Dirichlet process.

@@

In LDA, each document is represented by a mixture of the topics, with weight 
$\theta_w^2$ for topic $z$ in document $w$. These weight have a Dirichlet prior, 
parameterized by a hyper-parameter $\alpha$. Each topic is a probability 
distribution over the vocabulary, with probability $\beta^w_z$ for word $w$ in 
topic $z$.

For each document $w$, in corpus $D$, the LDA generative process is as follows:

\begin{enumerate}
	\item Draw topic distribution  $\theta_w \sim Dir(\alpha)$
	\item For each of the $N$ words $w_n$:
	\begin{description}
	\item (a) Draw a specific topic $z_n \sim Multinomial(\theta)$
	\item (b) Draw a word $w_n \sim Multinomial(\beta_{zn})$
	\end{description}
\end{enumerate}

As seen in step 1, each document contains topics in different proportions, determined 
by the scaling parameter $\alpha$. The individual words are drawn from one of the $k$ 
topics in step (b), where the probability of each word within a topic is parameterized 
by a $k \times V$ matrix , where $V$ is the size of the specified vocabulary. The 
most probable words in each topic could be used to identify the topic and gives an 
intuitive description for the topic.

The key difference in LDA compared to PLSI is that the process outlined above actually 
is generative for documents. This means that, given a new document, one can use the 
parameters learned previously to estimate the topic distribution $\theta$ for this 
new document.

A limitation with the standard LDA model is the fix vocabulary. This means that new 
words cannot enter the model and that you have to know a priori what words will model 
the corpus effectively. A setting where this could be problematic is when analyzing 
twitter data where new words in the form of slang, names and hashtags emerge over 
time. Another limitation is the fact that one has to state the number of topics 
the model should use. This is not a big issue, but could lead to some of the 
resulting topics to be insignificant.

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}

@@

The Expectation Maximization, or EM, algorithm is an approach to deriving an algorithm 
for approximately obtaining maximum likelihood of variables in a statistical model. 
Normally one obtains a maximum likelihood solution by taking partial derivatives of 
the likelihood function with respect to all variables, setting these derivatives to 
zero and solving the equations simultaneously.

Due to latent variables it is not always possible to solve the equations directly and 
instead one typically gets two sets of equations that depend on each other. The EM 
algorithm uses the idea that one could choose starting values for the latent variables 
in the first set of equations, use the result in the second set of equations, and then 
alternate between the two sets until convergence.

Consider a statistical model accompanied by a set of observed data $X$, a set of 
latent data $Z$, a set of latent variables $\theta$ and a likelihood function 
$L(\theta \mid X, Z)$. The marginal likelihood of the observed data is then

\begin{equation}
L(\theta \mid X) = p(X \mid \theta) = \sum\limits_{Z} p(X, Z \mid \theta)
\end{equation}

which determines the maximum likelihood estimate of the latent variables. The sum is 
however often intractable and the EM algorithm solves this problem by iterating the 
following two steps:

\begin{itemize}
\item \textbf{Expectation step (E - step)}: For the current estimate of the parameters 
$\theta^t$, calculate the expected value of the log likelihood function:
\begin{equation}
Q(\theta \mid \theta^t) \leftarrow \mathbb{E}_{Z \mid X, \theta^t} [\log(L(\theta \mid X, Z))]
\end{equation}
\item \textbf{Maximization step (M - step)} Maximize the expected value function 
with regard to the parameters:
\begin{equation}
\theta^{t + 1} \leftarrow arg \max\limits_{\theta} Q(\theta \mid \theta^t)
\end{equation}
\end{itemize}

@@

Variational inference can be seen as an extension of the EM algorithm and is used to 
approximate intractable integrals. The goal of variational inference is to approximate 
an intractable probability distribution $p$, with a tractable one $q$, in a way that 
makes them as close as possible given a similarity measure. In this thesis, 
variational inference is used to approximate the posterior distribution of latent 
variables.

For a bayesian model where x = x 1:n are observations and z = z 1:m are 
latent variables, the posterior distribution is

\begin{equation}
p(x \mid z) = \frac{p(x \mid z) p(z)}{\int\limits_{z} p(x \mid z)p(z)}
\end{equation}

For many models, this distribution is intractable due to the integral in the 
denominator. The posterior can be computed up to a constant, but the normalization 
constant cannot be computed. Variational inference methods can be used to approximate 
this integral. Applying variational inference to posterior distributions is sometimes 
called variational Bayes but is only an instance of a more general problem that 
variational inference solves.
